Live Functional Programming with Typed Holes

### 2018-02-20 discussion.tex
Section 2, wip

>>>>

\rkc{something about UI choices for displaying large indeterminate expressions}

\rkc{something about UI choices for displaying hole environment traces}

\rkc{something about patterns of use for debugging, such as inserting holes
around all expressions in tail position}

\rkc{\citet{Seidel2016} define a partial evaluation with a form of holes. compare.
compare ``flattened'' sumList indeterminate result to their trace. could be
useful to study whether evaluating the type-correct parts of the evaluation
trace help students focus on the part that has the error.}

### 2018-02-20 examples.tex
Section 2, wip

The process of writing new program fragments includes many states in which the
program under construction is, by definition, incomplete.
%
Furthermore, it is natural to ``jump back and forth'' between different places
in the program, iteratively and alternately developing the missing pieces.
%
The primary benefit of our approach is that such programs can be evaluated in
order to provide useful feedback during this editing workflow.

@@

\overviewExample{1}{Computing Weighted Averages}
%
The teacher begins writing a function called \li{weighted_average} that will
compute a weighted average, of type \li{float}, for each \li{student_data}
record.
%
The body of \li{weighted_average} is not yet written---marked with the
\emph{empty hole} on line \rkc{XXX}--- but the teacher jumps ahead to map the
function over the \li{grades} list.

@@

\noindent
%
\HazelnutLive{} evaluates this incomplete program, showing its
\emph{indeterminate} result in the sidebar on the right half of
\autoref{fig:grades-example}; this value is a list whose length is the same as
\li{grades}, where each element is the (indeterminate) result of evaluating the
hole expression inside \li{weighted_average}.
%
Notice that the hole on line \rkc{XXX} is assigned a unique static identifier
(\ie{}~\li{a}) and values it produces are assigned corresponding unique dynamic
identifiers (\ie{}~\li{a.1}, \li{a.2}, etc{}).
%
In contrast, the typical ad-hoc approach to simulating hole expressions, namely,
using placeholder expressions like \li{raise "Not yet implemented."}, would halt
the program as soon it is evaluated (in this case, inside the definition of
\li{List.map} when the function is applied to the first element of \li{grades}.
%
By ``evaluating around'' the hole in \HazelnutLive{}, the programmer can at
least see that the resulting expression does indeed evaluate to a list with the
correct number of (indeterminate) values.

Now the teacher returns to \li{weighted_average} to work on the missing
expression, with the goal of computing a weighted sum of each of a student's
grades.
%
The teacher builds the summand weighting the \li{hw1} score to account for 10\%
of the weighted average; the rest of the computation is a new hole expression
(\li{??_b}), yet to be filled in.

@@

\overviewExample{2}{Assigning Letter Grades}
%
The teacher's next task is to map the weighted averages to the letter grades A
through F (we will consider only ``whole'' letter grades, for simplicity).
%
The \li{grade_cutoffs} record type describes the minimum cutoff for each of
these six possible.
%
Initially, each of values in the \li{cutoffs} is a hole.

@@

\noindent
%
%% initially all holes, because it will be different year-to-year based on the %
%data, differences in course difficulty, and to satisfy fairness criteria.
%
Before starting to fill in the cutoffs, the teacher jumps ahead to write a
function \li{letter_grade} that will make the connection between \li{cutoffs}
and \li{weighted_averages}.
%
The teacher sorts the \li{weighted_averages} because she intends to look at the
data to help select the cutoff values, and then maps them to \li{letter_grades}.

@@

\noindent
%
When \HazelnutLive{} runs this program, the guard of the outermost conditional,
\li{n >= cutoffs.a}, is indeterminate because \li{cutoffs.a} is.
%
Therefore, each of the indeterminate expressions in \li{letter_grades} is the
entire expression body, albeit with different bindings for \li{n}.
%
Displaying such ``large'' indeterminate expressions can quickly consume all
available screen space, overwhelming the user with too much information.
%
\autoref{fig:XXX} shows how \HazelnutLive{} renders large indeterminate
expressions (\ie{}~anything other than ``small'' indeterminate value (a constant
or hole) or list of small values) simply as \li{..} to save space.
%
Hovering over this abbreviation (as shown in \autoref{fig:XXX} displays the full
indeterminate expression as a tooltip.
%
\autoref{sec:discussion} discusses this and other user interface concerns when
trying to display useful live feedback without overwhelming the user.
%
These usability factors are beyond the scope of our work, which is to define
semantic foundations on which such user interfaces can be built.

@@

\noindent
%
Running \li{compute_distribution} shows that the percentage of As is
\rkc{XXX\%}, which is smaller than what the teacher would like.
%
Returning to the value of \li{sorted_data}, she sees a cluster around \li{89.0}
and then another gap between \li{88.2} and {85.5}.
%
So, the teacher adjusts \li{cutoffs.a} to be \li{88.0}.
%
Because this edit is outside of a hole expression, \HazelnutLive{} discards the
previous execution state and reevaluates the entire program.
%
Existing techniques for incremental computation~\cite{XXX,XXX}, however, could
be applied to seek opportunities for reuse even when non-hole expressions are
modified.
%
Because the focus of work is on the novel implications of running programs with
holes, our calculus and implementation supports caching ``edit-and-resume'' only
for the novel situation in which evaluation proceeds around holes.
%
After re-evaluation, the percentage of As becomes \rkc{XXX\%}, which better
matches the teacher's intention.

@@

@@

Understanding and debugging static type errors is notoriously difficult,
particularly for novices.
%
A variety of approaches have been
proposed~\cite{Seminal,ChenErwig2014,Pavlinovic2015,sherrloc} to better localize
and explain type errors.
%
One of these approaches~\cite{Seidel2016} proposes to generate a dynamic witness
that demonstrates a run-time failure, and then displays a compressed execution
trace to the user as a graph.
%
By running type-incorrect programs as far as possible (in particular, until they
would go wrong), our semantics can provide a similar benefit operationally.
%
\rkc{they define a partial evaluation with a form of holes. compare.}

\emph{Example 3: Sum List.}
%
Consider the following buggy program of numbers reported by \citet{XXX} that
attempts to sum a list numbers.
%
The problem is that the base case produces a list rather than a number (in
particular, 0).

\noindent
%
Because the list expression on line 2 does not have type \li{int} as required,
it is wrapped in a (non-empty) hole.
%
Rather than trying to debug the error based on the static error, the programmer
may wish to run a call to the function, say \li{sumList(2)}, in which case
\HazelnutLive{} runs and produces the indeterminate value \li{1 + 2 + ?[]?}
(\cf{} the trace displayed by \citet{Seidel2016}).
%
By observing that the hole expression is being added to the number \li{2}, he
realizes that it needs to be a number, specifically, \li{0}.

\parasection{Example 4: Stutter}
%
As another example, consider the following function which attempts to produce a
list where every element is repeated twice (borrowed from \citep{XXX}).
%
The combiner function to \li{List.foldr} needs to produce a \li{list(int)}, but
it produces a \li{list(list(int)} instead.

\noindent
%
The editor has a choice about which expression to ``blame'' for the error; the
entire application that forms the body of the lambda is analyzed against the
return type \li{list(int)}, so that is a reasonable choice for the editor to
make; another would be to assume that the arguments \li{[x,x]} and \li{acc} are
both as intended and that only the function \li{(:)} is type-consistent.
%
Although one could imagine a setting in which a user would perform this
reasoning, let's assume the simplest approach for marking consistencies that
wraps the entire application.
%
Running this on \li{stutter [1,2,3]} produces the indeterminate value
\li{?  [1,1] : (? [2,2] : (? [3:3] : [] ?) ?) ?}, which shows the unfolding of
\li{List.foldr}.
%
Inside the innermost hole, the value is shown to be \li{[[3,3]]}; the user
realizes that there are too many levels of nesting.
%
He tries append in place of cons, which addresses the type inconsistency and
produces the desired result.

@@

\parasection{Example 5: \rkc{...}}

@@

\parasection{Example 6: \rkc{...}}

@@

>>>>

The process of writing new program fragments includes many states in which the
program under construction is, by definition, incomplete.
%
It is natural for a programmer to ``jump back and forth'' between different
places in the program, iteratively and alternately developing the missing
pieces.
%
The primary benefit of our approach is that such programs can be evaluated in
order to provide useful feedback during this editing workflow.

@@

\overviewExample{1}{Computing Weighted Averages}
%
The teacher begins writing a \li{weighted_average} function that will
compute a weighted average, of type \li{float}, for each \li{student_data}
record.
%
The body of \li{weighted_average} is not yet written---marked with the
\emph{empty hole} on line \rkc{XXX}--- but the teacher jumps ahead to map the
function over the \li{grades} list.

@@

\noindent
%
\HazelnutLive{} evaluates this incomplete program, showing its
\emph{indeterminate} result in the sidebar on the right half of
\autoref{fig:grades-example}; this value is a list whose length is the same as
\li{grades}, where each element is the (indeterminate) result of evaluating the
hole expression inside \li{weighted_average}.
%
Notice that the hole on line \rkc{XXX} is assigned a unique static identifier
(\ie{}~\li{a}) and that the indeterminate results it produces are assigned
corresponding unique dynamic identifiers (\ie{}~\li{a.1}, \li{a.2}, etc.).
%
In contrast, the typical ad-hoc approach to simulating hole expressions, namely,
using placeholder expressions like \li{raise "Not yet implemented."}, would halt
the program as soon it is evaluated (in this case, inside the definition of
\li{List.map} when the function is applied to the first element of \li{grades}.
%
By ``evaluating around'' the hole in \HazelnutLive{}, the programmer can at
least see that the resulting expression does indeed evaluate to a list with the
correct number of (indeterminate) values.

Now the teacher returns to \li{weighted_average} to work on the missing
expression, with the goal of computing a weighted sum of each of a student's
grades.
%
The teacher uses a \HazelnutLive{} edit action to create a sum expression,
filling in the first summand by weighting the \li{hw1} score to account for 10\%
of the weighted average; the rest of the computation is a new hole expression
(\li{??_b}), yet to be filled in.

@@

\overviewExample{2}{Assigning Letter Grades}
%
The teacher's next task is to map the weighted averages to the letter grades A
through F (we will consider only ``whole'' letter grades, for simplicity).
%
The \li{grade_cutoffs} record type describes the minimum cutoff for each of
these six possible grades.
%
Initially, each value in the \li{cutoffs} record is a hole.

@@

\noindent
%
%% initially all holes, because it will be different year-to-year based on the %
%data, differences in course difficulty, and to satisfy fairness criteria.
%
Before starting to fill in the cutoffs, the teacher jumps ahead to write a
function \li{letter_grade} that will make the connection between \li{cutoffs}
and \li{weighted_averages}.
%
Because she intends to look at the data to help select the cutoff values, the
teacher sorts the \li{weighted_averages} and then maps them to
\li{letter_grades}.

@@

\noindent
%
When \HazelnutLive{} runs this program, the guard of the outermost conditional
(\li{n >= cutoffs.a} on line \rkc{XXX}), is indeterminate because \li{cutoffs.a}
is.
%
Therefore, each of the indeterminate expressions in \li{letter_grades} is the
entire expression body, albeit with different bindings for \li{n}.
%
Displaying such ``large'' indeterminate expressions can quickly consume all
available screen space, overwhelming the user with too much information.
%
\autoref{fig:XXX} shows how \HazelnutLive{} renders large indeterminate
expressions (\ie{}~anything other than ``small'' indeterminate value (a constant
or hole) or list of small values) simply as \li{..} to save space.
%
Hovering over this abbreviation (as shown in \autoref{fig:XXX} displays the full
indeterminate expression---as well as the evaluation environment that surrounds
it---as a tooltip.
%
\autoref{sec:discussion} discusses this and other user interface concerns when
trying to display useful live feedback without overwhelming the user.
%
These usability factors are beyond the scope of our work, which is to define
semantic foundations on which such user interfaces can be built.

@@

\noindent
%
Running \li{compute_distribution} shows that the percentage of As is
\rkc{XXX\%}, which is smaller than what the teacher would like;
the remaining percentages are all indeterminate.
%
Returning to the value of \li{sorted_data}, she sees a cluster around \li{89.0}
and then another gap between \li{88.2} and {85.5}.
%
So, the teacher adjusts \li{cutoffs.a} to be \li{88.0}.
%
Because this edit is outside of a hole expression, \HazelnutLive{} discards the
previous execution state and reevaluates the entire program.
%
Existing techniques for incremental computation~\cite{XXX,XXX}, however, could
be applied to seek opportunities for reuse even when non-hole expressions are
modified.
%
Because the focus of work is on the novel implications of running programs with
holes, our calculus and implementation supports caching ``edit-and-resume'' only
for the novel situation in which evaluation proceeds around holes.
%
After re-evaluation, the percentage of As becomes \rkc{XXX\%}, which better
matches the teacher's intention.

@@

In the previous section, we described example programs that were incomplete
because of \emph{missing} expressions.
%
Next, we describe example programs that are incomplete because of
\emph{type-inconsistent} expressions.
%
For example, consider the following two definitions.

\begin{lstlisting}
let bad_bool : bool = ?? 0 ??_bad_bool;

let bad_int : int = 1 + ?? true ??_bad_int_second_argument;
\end{lstlisting}

\noindent
%
These two definitions are ill-typed under standard typing disciplines.
%
In contrast, \citet{popl-paper} present a bidirectional type system that assigns
types to both, by wrapping type-inconsistent expressions (\li{0} on line
\rkc{XXX} and \li{true} on line \rkc{XXX} above) in \emph{non-empty} holes.
%
Non-empty holes prevent local type inconsistencies from polluting the rest of
the program surrounding it, which may or may not itself contain additional
inconsistencies.

Understanding and debugging static type errors is notoriously difficult,
particularly for novices.
%
A variety of approaches have been
proposed~\cite{Seminal,ChenErwig2014,Pavlinovic2015,sherrloc} to better localize
and explain type errors.
%
One of these approaches~\cite{Seidel2016} proposes to generate a dynamic witness
that demonstrates a run-time failure, and then displays a compressed execution
trace to the user as a graph.
%
In \HazelnutLive{}, we can run programs with type errors (\ie{}, programs with
non-empty holes) as far as possible, in particular, until they
would go wrong.
%
Thus, our operational semantics provides a similar benefit to the approach of
\citet{Seidel2016}.

\overviewExample{3}{Sum List}
%
Consider the following buggy program (observed during an undergraduate
functional programming course~\cite{Seidel2016}) that attempts to sum a list
integers.
%
The error is that the base case produces a list rather than an integer.

\noindent
%
Because the list expression on line 2 does not have type \li{int} as required,
it is wrapped in a (non-empty) hole by the bidirectional type
checker~\cite{popl-paper}.
%
Rather than trying to debug the error based on the static error, the programmer
may wish to trying running the function anyway by calling, say, \li{sumList(2)}.
%
\HazelnutLive{} runs and produces the indeterminate expression \li{3 + ?[]?}.
%
By observing that the hole expression is being added to the integer \li{3}, he
realizes that it needs to be an integer, specifically, \li{0}.
%
Compared to the trace displayed by \citet{Seidel2016}, the indeterminate result
produced by \HazelnutLive{} is ``flattened'' because the expression \li{1 + 2}
successfully proceeded to evaluate despite the error elsewhere.

\overviewExample{4}{Stutter}
%
Consider the following function which attempts to produce a
list where every element is repeated twice (borrowed from \citet{Osera2015}).
%
The combiner function to \li{List.foldr} needs to produce a \li{list(int)}, but
it produces a \li{list(list(int)} instead.

\noindent
%
The bidirectional type checker of \citet{popl-paper} wraps the expression
\li{[x,x]} inside a non-empty hole.
%
%% The editor has a choice about which expression to ``blame'' for the error; the
%% entire application that forms the body of the lambda is analyzed against the
%% return type \li{list(int)}, so that is a reasonable choice for the editor to
%% make; another would be to assume that the arguments \li{[x,x]} and \li{acc} are
%% both as intended and that only the function \li{(:)} is type-consistent.
%% %
%% Although one could imagine a setting in which a user would perform this
%% reasoning, let's assume the simplest approach for marking consistencies that
%% wraps the entire application.
%
Running this on \li{stutter [1,2,3]} produces the indeterminate result

\begin{lstlisting}
?  [1,1] : (? [2,2] : (? [[3:3]] ?) ?) ?,
\end{lstlisting}

\noindent
%
which shows the unfolding of \li{List.foldr}.
%
We refer to nested indeterminate computations like this as \rkc{\emph{hole
environment traces} or \emph{hole environment trees}}.
%
The result of the innermost indeterminate expression is \li{[[3,3]]}.
%
The user realizes that there are too many levels of nesting, so
he replaces the \li{(:)} with \li{(@)}, which addresses the type inconsistency
and, when reevaluated, produces the desired result.

@@

@@

\overviewExample{5}{\rkc{...}}

@@

\overviewExample{6}{\rkc{...}}

@@

\paragraph{Recap}
%
In the above four subsections, respectively, we considered programs with:
%
(1) empty expression holes,
%
(2) non-empty expression holes for ill-typed expressions,
%
(3) type holes, and
%
(4) non-empty expression holes for well-typed expressions.
%
Next, we will formally describe the novel dynamic semantics of \HazelnutLive{}
that allows programs with combinations of these kinds of incompleteness to be
evaluated.
